Local wind regime induced by giant linear dunes: comparison of ERA5-Land reanalysis with surface measurements

### 2022-07-04 Paper/main.tex
Philippe round of review

\title{Local wind regime induced by giant linear dunes: comparison of ERA5-Land re-analysis with surface measurements}

@@

\begin{abstract}
Emergence and growth of sand dunes results from the dynamic interaction between topography, wind flow and sediment transport. While feedbacks between these variables are well studied at the scale of a single dune, the average effect of a periodic dune pattern on atmospheric flows remains poorly constrained due to a lack of data in major sand seas. Here, we compare wind tower measurements of surface wind data to the predictions of the ERA5-Land climate reanalysis at four locations in Namibia, within and outside the giant-dune field of the Namib sand sea. In the desert plains to the north of the sand sea, observations and predictions agree well. This is also the case in the interdune areas of the sand sea during the day, but not during the night, where additional components aligned with the giant dune orientation are measured instead of the easterly wind predicted by the ERA5-Land reanalysis.
%
These discrepancies, with wind deviation and attenuation larger then $50^\circ$ and $60~\%$, are linked to the daily cycle of the turbulent atmospheric boundary layer over a complex topography, and to the associated flow regimes. During the night, the low wind velocities within a shallow boundary layer induces a flow confinement, leading to a strong streamline compression above the giant dunes and thus resulting in large flow deviations. During, the day, the flow confinement is reduced by higher wind velocities and deep boundary layers.
%
We finally propose that, in multidirectional wind regimes, this mechanism and the resulting wind deflections by giant dunes could explain the occurrence of secondary dune patterns elongating in different orientations compared to the primary structures between which they develop.

@@

@@

Compared to mountains, aeolian sand dunes offer idealized topographies to the study of atmospheric turbulent flow over wavy bottoms, due to their smooth shape, free of canopies, resulting from their interaction with the wind \citep{Bagnold1941}. Then, past studies have highlighted two related topographic feedbacks on the windflow close to the dune/hill surface.
%
First is the effect on wind speed, with documented flow acceleration on upwind slopes \citep{Weaver2011} and deceleration on downwind slopes \citep{Baddock2007}, where the speed-up factor is essentially proportional to the obstacle aspect ratio \citep{Jackson1975}. Importantly, the velocity maximum is typically shifted upwind of the obstacle crest \citep{Jackson1975, Claudin2013}. This behaviour has been theoretically predicted by means of asymptotic analysis of a neutrally stratified boundary-layer flow over an obstacle of vanishing aspect ratio \citep{Jackson1975, Mason1979, Sykes1980, Hunt1988, Belcher1998}. Experiments in flumes \citep{Zilker1977, Zilker1979, Frederick1988, Poggi2007, Bristow2022}, in wind tunnels \citep{Gong1989, Finnigan1990, Gong1996} and in field conditions at all scales \citep{Taylor1987a, Claudin2013, Fernando2019, Lu2021}, have also documented this effect. Interestingly, a similar behaviour exists for the pressure perturbation, but with a slight downwind shift for the pressure minimum \citep{Claudin2021}. The second effect, much less studied, is the flow deflection that occurs when the incident wind direction is not perpendicular to the ridge crest. While predicted to be small (less than $10^{\circ}$) in the linear regime valid for shallow topography \citep{Gadal2019}, significant flow steering has been reported in the field on the downwind side of steep enough obstacles, such as well-developed sand dunes \citep{Tsoar1983, Sweet1990, Walker2002, Smith2017} and in particular coastal foredunes (e.g.~\citet{rasmussen1989}, \citet{Walker2006}, \citet{Walker2009}, \citet{Hesp2015}, \citet{Walker2017}, \citet{deWinter2020}), mountain ranges \citep{Kim2000, Lewis2008, Fernando2019}, and valley topographies \citep{Wiggs2002, Garvey2005}.

Wind measurements over sand dunes has been mainly performed over small bedforms, typically a few meters high (corresponding to tens of meters long) (e.g.~\citet{Mulligan1988}, \citet{Hesp1989}, \citet{Lancaster1996}, \citet{MckennaNeuman1997}, \citet{Sauermann2003}, \citet{Andreotti2002}, \citet{Walker2002}, \citet{Weaver2011}). For practical reasons, fewer studies performed similar measurements on giant dunes \citep{Havholm1988}, with kilometer-scale wavelengths and heights of tens of meters. However, they provide a choice configuration for the study of turbulent flows over a complex topography. First, one expects larger wind disturbances for larger obstacles. Secondly, their large size makes them interact with the vertical structure of the atmosphere \citep{Andreotti2009}. Third, they usually form large patterns in sand seas and thus behave as rather clean periodic perturbations, in contrast with isolated dunes. Finally, because the morphodynamics of aeolian bedforms are strongly dependent on the local wind regime \citep{Livingstone2019}, one can expect to see the consequences of windflow disturbance by large dunes on neighbouring small dunes \citep{Brookfield1977, Ewing2006}. A similar effect is observed on the properties of impact ripple patterns due to the presence of dunes \citep{Howard1977, Hood2021}.

Atmospheric flows have been much studied at the desert-scale with climate reanalyses based on global atmospheric models \citep{Blumberg1996, Livingstone2010, Ashkenazy2012, Jolivet2021, Hu2021}, such as ERA-40, ERA-Interim or ERA-5 \citep{Uppala2005, Dee2011, Hersbach2020}. However, the spatial resolution (tens of kilometers) of these reanalyses implies average quantities that do not resolve the smaller scales of interest, which range from individual dunes to small mountains \citep{Livingstone2010}. Recently, the release of ERA5-Land has resolved this limitation by providing up to 70 years of hourly wind predictions at a $9$~km spatial resolution \citep{munoz2021}. However, its validity remains to be studied, especially in remote desert areas where assimilation of measured data is very low.

In this work, we compare local wind speeds and directions measured by meteorological stations at four different locations inside and north of the giant-dune field of the Namib sand sea to the regional predictions of the ERA5-Land climate reanalysis. Where the meteorological stations are surrounded by a relatively flat environment, we show that local measurements and regional predictions agree well. The agreement is also good in the interdune areas of the sand sea, except for some weak winds blowing at night, which exhibit an additional component aligned with the giant dune orientation. These winds are not predicted by the ERA5-Land reanalysis (section 2). Further, we are able to link the magnitude of these differences to the circadian cycle of the atmospheric boundary layer (section 3). Finally, we draw implications for the wind disturbances on smaller-scale dunes (section 4), suggesting a possible origin for crossing dunes.

@@

%---------------------------------------------
\begin{figure}[t]
\centering
\includegraphics[scale=1]{Figures/Figure2.pdf}
\caption{Wind data used in this study. Satellite images of these different environements (Google-Earth, Maxar Technologies, CNES/Airbus) are shown on the left. The black dots show the location of the wind measurements stations. On the center and on the right, wind roses representing the data from the ERA5-Land climate reanalysis and the local wind stations are respectively shown. Note that the bars show the direction towards which the wind blows.}
\label{Fig2}
\end{figure}
%---------------------------------------------

@@

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure3.pdf}
\caption{Temporal comparison between the wind data coming from the ERA5-Land climate reanalysis (orange lines) and from the local measurements (blue lines). Coloured swathes indicate day (between 1000 UTC and 2200 UTC) and night (before 1000 UTC or after 2200 UTC). Numbers indicate the average flow deflection $\delta_{\theta}$ and wind modulation $\delta_{u}$ over the given period, as well as the frequency of occurence of extreme events ($\delta_{\theta} > 50^\circ$, $\vert\delta_{u}\vert > 0.6$). \textbf{a--b}: Huab station in summer. \textbf{b--c}: Huab station in winter. \textbf{d--e}: Etosha West station in summer. \textbf{f--g}: Etosha West station in winter. Time series of the two other stations are shown in Fig.~\ref{Fig5}.}
\label{Fig3}
\end{figure}
%---------------------------------------------

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure4.pdf}
\caption{Distributions of wind direction at the Huab and Etosha West stations for the
ERA5-Land climate reanalysis (orange) and the local measurements (blue). In each subplot,
both distributions are plotted from the same time steps, selected with constraints on the
wind wind velocity (rows) in the ERA5-Land dataset. The numbers at the top centre give
the percentage of time steps selected in each sub-range, as well as the percentage of them
corresponding to the day (between 1000 UTC and 2200 UTC).}
\label{Fig4}
\end{figure}
%---------------------------------------------

@@

In the case of the Etosha West and Huab stations, the time series of wind speed and direction from the regional predictions quantitatively match those corresponding to the local measurements (Figs.~\ref{Fig3}, ~\ref{Fig4} and Online Resource Fig.~\ref{Fig5_supp}). For the North Sand Sea and South Sand Sea stations within the giant dune field, we observe that this agreement is also good, but limited to the October-March time period (Fig.~\ref{Fig4}a, b, e, f). However, the field-measured wind roses exhibit additional wind components aligned with the giant dune orientation, as evidenced on the satellite images (Fig.~\ref{Fig2}c,d).

More precisely, during the April-September period, the local and regional winds in the interdune match during daytime only, i.e when the southerly/southwesterly sea breeze dominates (Figs.~\ref{Fig5}c,d,g,h and \ref{Fig6}). In the late afternoon and during the night, when the easterly `berg' and katabatic winds blow, measurements and predictions differ. In this case, the angular wind distribution of the local measurements exhibits two additional modes corresponding to reversing winds aligned with the giant dune orientation (purple frame in Fig.~\ref{Fig6}, Online Resource Fig.~\ref{Fig6_supp}). This deviation is also associated with a general attenuation of the wind strength (Online Resource Fig.~\ref{Fig7_supp}). Remarkably, all these figures show that these wind reorientation and attenuation processes occur only at low velocities of the regional wind, typically for $u_{*, \, \textup{ERA}} \lesssim 0.2~\textrm{m}~\textrm{s}^{-1}$. For shear velocities larger than $u_{*, \, \textup{ERA}} \simeq 0.3~\textrm{m}~\textrm{s}^{-1}$, the wind reorientation is not apparent. Finally, for intermediate shear velocities, both situations of wind flow reoriented along the dune crest and not reoriented can be successively observed (Online Resource Fig.~\ref{Fig6_supp}). Importantly, these values are not precise thresholds, but indicative of a crossover between regimes, whose physical interpretation is discussed in the next section.

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure5.pdf}
\caption{Temporal comparison between the wind data coming from the ERA5-Land climate reanalysis (orange lines) and from the local measurements (blue lines). Coloured swathes indicate day (between 1000 UTC and 2200 UTC) and night (before 1000 UTC or after 2200 UTC). Numbers indicate the average flow deflection $\delta_{\theta}$ and wind modulation $\delta_{u}$ over the given period, as well as the frequency of occurence of extreme events ($\delta_{\theta} > 50^\circ$, $\vert\delta_{u}\vert > 0.6$). \textbf{a--b}: North Sand Sea station in summer. \textbf{b--c}: North Sand Sea station in winter. \textbf{d--e}: South Sand Sea in summer. \textbf{f--g}: South Sand Sea station in winter. Time series of the two other stations are shown in Fig.~\ref{Fig3}.}
\label{Fig5}
\end{figure}
%---------------------------------------------

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure6.pdf}
\caption{Distributions of wind direction at the North Sand Sea (\textbf{a}) and South Sand Sea (\textbf{b}) stations for the ERA5-Land climate reanalysis (orange) and the local measurements (blue). In each subplot, both distributions are plotted from the same time steps, selected with constraints on the wind direction (columns) and/or wind velocity (rows) of the ERA5-Land dataset. The grey vertical dashed lines indicate the dune orientation. The numbers at the top right give the percentage of time steps selected in each sub-range, as well as the percentage corresponding to the daytime (between 1000 UTC and 2200 UTC). Contrary to the Huab and Etosha West stations (Fig.~\ref{Fig4}), histograms does not match low wind velocities. More specifically, the purple frame highlights the regime (low wind velocities, nocturnal easterly wind) in which the data from both datasets differ.}
\label{Fig6}
\end{figure}
%---------------------------------------------

@@

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure7.pdf}
\caption{\textbf{a--b}: Vertical profiles of the virtual potential temperature at 2 different time steps (day - 03/11/2015 - 1200 UTC, night - 01/13/2013 - 0900 UTC) at the North Sand Sea station. Dots: data from the ERA5 reanalysis. Dashed lines: boundary layer height given by the ERA5 reanalysis (Online Resource section 2). Plain lines: vertical (boundary layer) and linear (free atmosphere) fits to estimate the stratification properties. \textbf{c--f}: Sketches representing the interaction between the giant dunes and the atmospheric flow for different meteorological conditions. \textbf{g}: Streamlines over a sinusoidal topography $\xi(x,y)$ qualitatively representing the effect of low, medium and strong flow confinement, in relation to the above panels (see Appendix~1 for more details). The blue arrow indicates the undisturbed wind direction.}
\label{Fig7}
\end{figure}
%---------------------------------------------

@@

Flow confinement below a capping surface, which enhances streamline compression, has a strong effect on the hydrodynamic response and typically increases flow deflection. This is the case for bedforms forming in open channel flows such as rivers~\citep{Kennedy1963, Chang1970, Mizumura1995, Colombini2004, Fourriere2010, Andreotti2012, Unsworth2018}. This is also relevant for aeolian dunes as they evolve in the turbulent atmospheric boundary layer (ABL) capped by the stratified free atmosphere (FA) \citep{Andreotti2009}. Two main mechanisms, associated with dimensionless numbers must then be considered (Fig.~\ref{Fig7}). First, topographic obstacles typically disturb the flow over a characteristic height similar to their length. As flow confinement is characterised by a thickness $H$, the interaction between the dunes and the wind in the ABL is well captured by the parameter $k H$. The height $H$ is directly related to the radiative fluxes at the Earth surface. It is typically on the order of a kilometre, but significantly varies with the circadian and seasonal cycles. Emerging and small dunes, with wavelengths in the range $20$ to $100$~m, are not affected by the confinement, corresponding to $k H \gg 1$. For giant dunes with kilometric wavelengths, however, their interaction with the FA is significant \citep{Andreotti2009}. This translates into a parameter $kH$ in the range $0.02$--$5$, depending on the moment of the day and the season. A second important mechanism is associated with the existence of a thin intermediate so-called capping layer between the ABL and the FA. It is characterised by a density jump $\Delta\rho$, which controls the `rigidity' of this interface, i.e. how much its deformation affects streamline compression. This is usually quantified using the Froude number \citep{Vosper2004, Stull2006, Sheridan2006, Hunt2006, Jiang2014}:
%
\begin{equation}
\mathcal{F} = \displaystyle\frac{U}{\sqrt{\displaystyle\frac{\Delta\rho}{\rho_{0}}gH}},
\label{FroudeNumber}
\end{equation}
%
where $U$ is the wind velocity at the top of the ABL and $\rho_{0}$ its average density. The intensity of the stratification, i.e. the amplitude of the gradient $\left | \partial_z \rho \right|$, also impacts its ability to deform the capping layer under the presence of an underlying obstacle, and thus affects the influence of flow confinement. This can be quantified using the internal Froude number~\citep{Vosper2004, Stull2006, Sheridan2006, Hunt2006, Jiang2014} $\mathcal{F}_{\textup{I}} = k U/N$, where $N = \sqrt{-g \partial_z \rho / \rho_{0}}$ is the the Brunt-V\"ais\"al\"a frequency \citep{Stull1988}. Both Froude numbers have in practice the same qualitative effect on flow confinement, and we shall restrict the main discussion to $\mathcal{F}$ only.

@@

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure8.pdf}
\caption{Regime diagrams of the wind deviation $\delta_{\theta}$ (\textbf{a}) and relative attenuation/amplification $\delta_{u}$ (\textbf{b}) in the space $(\mathcal{F}, \, kH)$, containing the data from both the North Sand Sea and South Sand Sea stations. Green dashed lines empirically delimit the different regimes. The point density in each bin of the diagrams is shown in Online Resource Fig.~\ref{Fig10_supp} -- 95\% of the data occur in the range $-1 < \delta u < 1$. Similar regime diagrams in the spaces $(\mathcal{F}_{\textup{I}}, \, kH)$ and $(\mathcal{F}_{\textup{I}}, \, \mathcal{F})$ are shown in Online Resource Fig.~\ref{Fig11_supp}.}
\label{Fig8}
\end{figure}
%---------------------------------------------

>>>>

\title{Local wind regime induced by giant linear dunes: comparison of ERA5-Land reanalysis with surface measurements}

@@

\begin{abstract}
Emergence and growth of sand dunes results from the dynamic interaction between topography, wind flow and sediment transport. While feedbacks between these variables are well studied at the scale of a single and relatively small dune, the average effect of a periodic large-scale dune pattern on atmospheric flows remains poorly constrained, due to a lack of data in major sand seas. Here, we compare local measurements of surface winds to the predictions of the ERA5-Land climate reanalysis at four locations in Namibia, within and outside the giant linear dune field of the Namib sand sea. In the desert plains to the north of the sand sea, observations and predictions agree well. This is also the case in the interdune areas of the sand sea during the day. During the night, however, an additional wind component aligned with the giant dune orientation is measured, in contrast to the easterly wind predicted by the ERA5-Land reanalysis.
%
We link these discrepancies, with wind deviation and velocity attenuation larger then $50^\circ$ and $60~\%$, to the daily cycle of the turbulent atmospheric boundary layer over a complex topography, and to the associated flow regimes. During the night, a shallow boundary layer induces a flow confinement associated with a strong streamline compression above the giant dunes, resulting in large flow deviations, especially for the the lower winds. During the day, the flow confinement is reduced by a thicker boundary layer and higher wind velocities, and the feedback of the giant dunes on the atmospheric flow is negligible.
%
We finally propose that this mechanism and the resulting wind deflections by the giant dunes could explain the occurrence of smaller-scale secondary dune patterns, elongating along a different orientation compared to the primary dunes between which they develop.

@@

CITE \citet{Chand22}

@@

Compared to hills and mountains, aeolian sand dunes offer idealized elevation profiles for the study of atmospheric turbulent flow over topographies, due to their smooth shape, free of canopies. Besides, dunes provide a rather wide range of scales, from decameters to kilometers, and very often come in a fairly regular pattern, which further simplifies the flow structure analysis. Past studies have highlighted two important topographic feedbacks on the wind flow close to the dune/hill surface. First is the effect on wind speed, with documented flow acceleration on upwind slopes \citep{Weaver2011} and deceleration on downwind slopes \citep{Baddock2007}, where the speed-up factor is essentially proportional to the obstacle aspect ratio \citep{Jackson1975}. Importantly, the velocity maximum is typically shifted upwind of the obstacle crest \citep{Jackson1975, Claudin2013}. This behaviour has been theoretically predicted by means of asymptotic analysis of a neutrally stratified boundary-layer flow over an obstacle of vanishing aspect ratio \citep{Jackson1975, Mason1979, Sykes1980, Hunt1988, Belcher1998, Kroy2002}. Experiments in flumes \citep{Zilker1977, Zilker1979, Frederick1988, Poggi2007, Bristow2022}, in wind tunnels \citep{Gong1989, Finnigan1990, Gong1996} and in field conditions at all scales \citep{Taylor1987a, Claudin2013, Fernando2019, Lu2021}, have also documented this effect. Interestingly, a similar behaviour exists for the pressure perturbation, but with a slight downwind shift for the pressure minimum \citep{Claudin2021}. The second effect, much less studied, is the flow deflection that occurs when the incident wind direction is not perpendicular to the ridge crest. While predicted to be small (less than $10^{\circ}$) in the linear regime valid for shallow topography \citep{Gadal2019}, significant flow steering has been reported in the field on the downwind side of steep enough obstacles, such as well-developed sand dunes \citep{Tsoar1983, Sweet1990, Walker2002, Smith2017} and in particular coastal foredunes (e.g.~ \citet{Hunter1983}, \citet{rasmussen1989}, \citet{Walker2006}, \citet{Walker2009}, \citet{Hesp2015}, \citet{Walker2017}, \citet{deWinter2020}), mountain ranges \citep{Kim2000, Lewis2008, Fernando2019}, and valley topographies \citep{Wiggs2002, Garvey2005}.

Wind measurements over sand dunes has been mainly performed over small bedforms, typically a few meters high (corresponding to several tens of meters long) (e.g.~\citet{Mulligan1988}, \citet{Hesp1989}, \citet{Lancaster1996}, \citet{MckennaNeuman1997}, \citet{Sauermann2003}, \citet{Andreotti2002}, \citet{Walker2002}, \citet{Weaver2011}). For practical reasons, fewer studies performed similar measurements on giant dunes \citep{Havholm1988}, with kilometer-scale wavelengths and heights of tens of meters. However, such large dunes provide a choice configuration for the study of turbulent flows over a complex topography. First, one expects larger wind disturbances for larger obstacles. Secondly, their large size can make them interact with the vertical structure of the atmosphere \citep{Andreotti2009}. Third, they usually form large patterns in sand seas and thus behave as rather clean periodic perturbations, in contrast with isolated dunes. Finally, because the morphodynamics of aeolian bedforms is strongly dependent on the local wind regime \citep{Livingstone2019}, one can expect to see the consequences of windflow disturbance by large dunes on neighbouring small dunes \citep{Brookfield1977, Ewing2006}. A similar effect is observed on the properties of impact ripple patterns due to the presence of dunes \citep{Howard1977, Hood2021}.

Atmospheric flows have been much studied at the desert-scale with climate reanalyses based on global atmospheric models \citep{Blumberg1996, Livingstone2010, Ashkenazy2012, Jolivet2021, Hu2021, gunn2021}, such as ERA-40, ERA-Interim or ERA5 \citep{Uppala2005, Dee2011, Hersbach2020}. However, the spatial resolution of these reanalyses (tens of kilometers) implies average quantities that do not resolve the smaller scales of interest, which range from individual dunes to small mountains \citep{Livingstone2010}. Recently, the release of ERA5-Land has partly resolved this limitation by providing up to 70 years of hourly wind predictions at a $9$~km spatial resolution \citep{munoz2021}. However, its validity remains to be studied, especially in remote desert areas where assimilation of measured data is very low.

In this work, we compare local wind speeds and directions measured by meteorological stations at four different locations inside and north of the giant linear dune field of the Namib sand sea to the regional predictions of the ERA5-Land climate reanalysis. Where the meteorological stations are surrounded by a relatively flat environment, we show that local measurements and regional predictions agree well. The agreement is also good in the interdune areas of the sand sea, except for some weak winds blowing at night, which exhibit an additional component aligned with the giant dune orientation. These winds are not predicted by the ERA5-Land reanalysis (section 2). Further, we are able to link the magnitude of these differences to the circadian cycle of the atmospheric boundary layer (section 3). Finally, we draw implications for the wind disturbances on smaller-scale dunes (section 4), suggesting a possible origin for crossing dunes.

@@

%---------------------------------------------
\begin{figure}[t]
\centering
\includegraphics[scale=1]{Figures/Figure2.pdf}
\caption{Wind data used in this study. Satellite images of the different environments (Google-Earth, Maxar Technologies, CNES/Airbus) are shown on the left. The black dots show the location of the wind measurements stations. On the right of the photos, the corresponding wind roses representing the data from the ERA5-Land climate reanalysis and the local wind stations are displayed. Note: the graphical convention for the wind roses is that the bars show the direction towards which the wind blows (see color bar for velocity scale).}
\label{Fig2}
\end{figure}
%---------------------------------------------

@@

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure3.pdf}
\caption{Temporal comparison between the wind data coming from the ERA5-Land climate reanalysis (orange lines) and from the local measurements (blue lines). Coloured swathes indicate day (between 10.00 UTC and 22.00 UTC) and night (before 10.00 UTC or after 22.00 UTC). Numbers in legends indicate the average flow deflection $\delta_{\theta}$ and relative wind modulation $\delta_{u}$ over the displayed period (see section~\ref{DataDistribution} for their definitions), as well as the percentage $f_\theta$ and $f_u$ of occurrence of extreme events ($\delta_{\theta} > 50^\circ$, $\vert\delta_{u}\vert > 0.6$). \textbf{a--b}: Huab station in summer. \textbf{b--c}: Huab station in winter. \textbf{d--e}: Etosha West station in summer. \textbf{f--g}: Etosha West station in winter. Time series of the two other stations are shown in Fig.~\ref{Fig5}.}
\label{Fig3}
\end{figure}
%---------------------------------------------

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure4.pdf}
\caption{Distributions of wind direction at Huab and Etosha West stations for the ERA5-Land climate reanalysis (orange) and the local measurements (blue). In each subplot, both distributions are plotted from the same time steps, selected for different ranges of the wind wind velocity (rows) in the ERA5-Land dataset. The numbers at the upper right corners give the percentage of time steps selected in each sub-range (top), as well as the percentage of them corresponding to the day -- defined between 10.00 UTC and 22.00 UTC (bottom).}
\label{Fig4}
\end{figure}
%---------------------------------------------

@@

In the case of the Etosha West and Huab stations, the time series of wind speed and direction from the regional predictions quantitatively match those corresponding to the local measurements (Figs.~\ref{Fig3}, ~\ref{Fig4} and Online Resource Fig.~\ref{Fig5_supp}). For the North Sand Sea and South Sand Sea stations within the giant linear dune field, we observe that this agreement is also good, but limited to the October-March time period (Fig.~\ref{Fig4}a, b, e, f). However, the field-measured wind roses exhibit additional wind components aligned with the dune orientation, as evidenced on the satellite images (Fig.~\ref{Fig2}c,d).

More precisely, during the April-September period, the local and regional winds in the interdune match during daytime only, i.e when the southerly/southwesterly sea breeze dominates (Figs.~\ref{Fig5}c,d,g,h and \ref{Fig6}). In the late afternoon and during the night, when the easterly `berg' and katabatic winds blow, measurements and predictions differ. In this case, the angular wind distribution of the local measurements exhibits two additional modes corresponding to reversing winds aligned with the dune orientation (purple frame in Fig.~\ref{Fig6}, Online Resource Fig.~\ref{Fig6_supp}). This deviation is also associated with a general attenuation of the wind strength (Online Resource Fig.~\ref{Fig7_supp}). Remarkably, all these figures show that these wind reorientation and attenuation processes occur only at low velocities of the regional wind, typically for $u_{*}^{\textup{ERA5-Land}} \lesssim 0.2~\textrm{m}~\textrm{s}^{-1}$. For shear velocities larger than $u_{*}^{\textup{ERA5-Land}} \simeq 0.3~\textrm{m}~\textrm{s}^{-1}$, the wind reorientation is not apparent. Finally, for intermediate shear velocities, both situations of wind flow reoriented along the dune crest and not reoriented can be successively observed (Online Resource Fig.~\ref{Fig6_supp}). Importantly, these values are not precise thresholds (and certainly not related to the threshold for sediment transport), but indicative of a crossover between regimes, whose physical interpretation is discussed in the next section.

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure5.pdf}
\caption{Same as Fig.~\ref{Fig3} for North Sand Sea station in summer (\textbf{a--b}), North Sand Sea station in winter (\textbf{b--c}), South Sand Sea station in summer (\textbf{d--e}) and South Sand Sea station in winter (\textbf{f--g}).}
\label{Fig5}
\end{figure}
%---------------------------------------------

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure6.pdf}
\caption{Same as Fig.~\ref{Fig4} but for North Sand Sea (\textbf{a}) and South Sand Sea (\textbf{b}) stations. Here, subplots correspond to different ranges for the wind direction (columns) and wind velocity (rows) of the ERA5-Land dataset. The grey vertical dashed lines indicate the main dune orientation. In contrast with observations at the Huab and Etosha West stations (Fig.~\ref{Fig4}), histograms do not match well at low wind velocities, and the purple frame highlights the regime (low wind velocities, nocturnal easterly wind) in which the data from both datasets differ most.}
\label{Fig6}
\end{figure}
%---------------------------------------------

@@

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure7.pdf}
\caption{\textbf{a--b}: Vertical profiles of the virtual potential temperature $T_{\rm vp}$ at two different time steps (day - 03/11/2015 - 12.00 UTC, night - 01/13/2013 - 09.00 UTC) at the North Sand Sea station. Dots: data from the ERA5 reanalysis. Dashed lines: boundary layer height given by the ERA5 reanalysis (Online Resource section 2). Plain lines: vertical (boundary layer) and linear (free atmosphere) fits to estimate the stratification properties. \textbf{c--f}: Sketches representing the interaction between the giant dunes and the atmospheric flow for different meteorological conditions. \textbf{g}: Streamlines over a sinusoidal topography $\xi(x,y)$ qualitatively representing the effect of low, medium and strong flow confinement, in relation to the above panels (see Appendix~1 for more details). The blue arrow indicates the undisturbed wind direction.}
\label{Fig7}
\end{figure}
%---------------------------------------------

@@

Flow confinement below a capping surface, which enhances streamline compression, has a strong effect on the hydrodynamic response and typically increases flow deflection. This is the case for bedforms forming in open channel flows such as rivers~\citep{Kennedy1963, Chang1970, Mizumura1995, Colombini2004, Fourriere2010, Andreotti2012, Unsworth2018}. This is also relevant for aeolian dunes as they evolve in the turbulent atmospheric boundary layer (ABL) capped by the stratified free atmosphere (FA) \citep{Andreotti2009}. Two main mechanisms, associated with dimensionless numbers must then be considered (Fig.~\ref{Fig7}). First, topographic obstacles typically disturb the flow over a characteristic height similar to their length. As flow confinement is characterised by a thickness $H$, the interaction between the dunes and the wind in the ABL is well captured by the parameter $k H$. The height $H$ is directly related to the sensitive heat flux from the Earth surface. It is typically on the order of a kilometre, but significantly varies with the circadian and seasonal cycles. Emerging and small dunes, with wavelengths in the range $20$ to $100$~m, are not affected by the flow confinement, corresponding to $k H \gg 1$. For giant dunes with kilometer-scale wavelengths, however, their interaction with the FA can be significant \citep{Andreotti2009}. This translates into a parameter $kH$ in the range $0.02$--$5$, depending on the moment of the day and the season. A second important mechanism is associated with the existence of a thin intermediate so-called capping layer between the ABL and the FA. It is characterised by a density jump $\Delta\rho$, which controls the `rigidity' of this interface, i.e. how much its deformation affects streamline compression. This is usually quantified using the Froude number \citep{Vosper2004, Stull2006, Sheridan2006, Hunt2006, Jiang2014}:
%
\begin{equation}
\mathcal{F} = \displaystyle\frac{U}{\sqrt{\displaystyle\frac{\Delta\rho}{\rho_{0}}gH}},
\label{FroudeNumber}
\end{equation}
%
where $U$ is the wind velocity at the top of the ABL and $\rho_{0}$ its average density. The intensity of the stratification, i.e. the amplitude of the gradient $\left | \partial_z \rho \right|$ in the FA, also impacts the ability to deform the capping layer under the presence of an underlying obstacle, and thus affects the influence of flow confinement. This can be quantified using the internal Froude number~\citep{Vosper2004, Stull2006, Sheridan2006, Hunt2006, Jiang2014} $\mathcal{F}_{\textup{I}} = k U/N$, where $N = \sqrt{-g \partial_z \rho / \rho_{0}}$ is the the Brunt-V\"ais\"al\"a frequency \citep{Stull1988}. Both Froude numbers have in practice the same qualitative effect on flow confinement (a smaller Froude corresponding to a stiffer interface), and we shall restrict the main discussion to $\mathcal{F}$ only.

@@

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure8.pdf}
\caption{Regime diagrams of the wind deviation $\delta_{\theta}$ (\textbf{a}) and relative attenuation/amplification $\delta_{u}$ (\textbf{b}) in the space $(\mathcal{F}, \, kH)$, containing the data from both the North Sand Sea and South Sand Sea stations. The green dashed lines empirically delimit the different regimes. The point density in each bin of the diagrams is shown in Online Resource Fig.~\ref{Fig10_supp} -- 95\% of the data occur in the range $-1 < \delta u < 1$. Similar regime diagrams in the spaces $(\mathcal{F}_{\textup{I}}, \, kH)$ and $(\mathcal{F}_{\textup{I}}, \, \mathcal{F})$ are shown in Online Resource Fig.~\ref{Fig11_supp}.}
\label{Fig8}
\end{figure}
%---------------------------------------------

%------------------------
\subsection{Data distribution in the flow regimes}
\label{DataDistribution}
We can go one step further and analyse how our data quantitatively spread over the different regimes discussed above. For that purpose, one needs to compute $kH$ and $\mathcal{F}$ from the time series. $H$, $U$ and the other atmospheric parameters can be deduced from the various vertical profiles (temperature, humidity) available in the ERA5 climate reanalysis (Online Resource section 2). We quantify the flow deflection $\delta_{\theta}$ as the minimal angle between the wind orientations comparing the local measurements and the regional predictions. We also compute the relative velocity modulation as
%
\begin{equation}
\delta_{\textup{u}} = \frac{u_{*}^{\textup{ERA5-Land}} -  u_{*}^{\textup{Local mes.}}}{u_{*}^{\textup{ERA5-Land}}}.
\end{equation}
%
These two quantities are represented as maps in the plane $(\mathcal{F}, \, kH)$ (Fig.~\ref{Fig8}a,b), and one can clearly identify different regions in these graphs. Small wind disturbances (small $\delta_{\theta}$ and $\delta_{\textup{u}}$) are located in the top-right part of the diagrams, corresponding to a regime with low-interaction as well as low-confinement ($k H$ and $\mathcal{F}$ large enough, Fig.~\ref{Fig7}d). Lower values of $k H$ (stronger interaction) or of Froude number (stronger confinement) both lead to an increase in wind disturbances, both in terms of orientation and velocity. Below a crossover value $k H \simeq 0.3$, wind disturbance is less sensitive to the $\mathcal{F}$-value. This is probably due to enhanced non-linear effects linked to flow modulation by the obstacle when confinement is strong (e.g. wakes and flow recirculations). The Froude number also controls a transition from damped to amplified wind velocities in the interdune, with a crossover around $\mathcal{F} \simeq 0.4$ (Fig.~\ref{Fig8}b). Such an amplification is rather unexpected. Checking the occurrence of the corresponding data, it appears that these amplifications are associated with the southerly sea breeze, and occur dominantly during the October-March period, when the other easterly wind is not present (Online Resource Fig.~\ref{Fig12_supp}a--b). Furthermore, they occur less frequently during the afternoon, and more frequently at the end of the day (Online Resource Fig.~\ref{Fig12_supp}c). This effect may be linked to a change in the flow behaviour in the lee side of the obstacles but further measurements are needed in order to assess the different possibilities \citep{baines1995, Vosper2004}.
